The Nuclear Norm and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-08-21 conf_article/neon.tex
Moved stochastic into the appendix

{\bf Stochastic gradient estimator.}
    We assume access to a stochastic estimator $g(\cdot\,;\xi)\colon \Rmn \to \Rmn$ of the gradient $\nabla f(\cdot)$, where $\xi \sim \cD$ is a random variable sampled from a probability distribution $\cD$. We assume that the stochastic gradient estimator $g(\cdot;\xi)$ is unbiased and has bounded variance, that is, the following relations hold:
    \begin{aequation}\label{eq:variance}
    \Ed{\xi \sim \cD}{g(\mX;\xi)} = \nabla f(\mX)
    \quad\text{and}\quad
    \Ed{\xi \sim \cD}{\sqnf{g(\mX;\xi) - \nabla f(\mX)}} \leq \sigma^2
    \quad\text{for all}\;
    \mX \in \Rmn,
    \end{aequation}
    where $\sigma > 0$ is a positive variance parameter, and $\normf{\cdot}$ is the standard Euclidean, i.e. Frobenius, norm induced by the inner product $\<\cdot,\cdot>$, i.e., $\normf{\mX} = \sqrt{\<\mX,\mX>}=\sqrt{\tr(\mX^\top \mX)}$. These assumptions have been widely adopted for the analysis of many stochastic gradient optimization algorithms \citep{ghadimi2013stochastic,ghadimi2016accelerated,cutkosky2020momentum,sun2023momentum,horvath2023stochastic,gorbunov2020linearly}.

{\bf Non-Euclidean norm setting and Lipschitz continuous gradient.}
    We assume that matrix space $\Rmn$ is equipped with a norm $\norm{\cdot}\colon \Rmn \to \R_+$, which possibly does not coincide with the Frobenius, norm $\normf{\cdot}$. In addition, we assume that the gradient $\nabla f(\cdot)$ is Lipschitz continuous with respect to the norm $\norm{\cdot}$, that is, the following inequality holds:
    \begin{aequation}\label{eq:L}
    \norm{\nabla f(\mX) - \nabla f(\mX')}^\dagger \leq L \norm{\mX-\mX'}
    \quad\text{for all}\;
    \mX, \mX' \in \Rmn,
    \end{aequation}
    where $L > 0$ is the gradient Lipschitz constant, and $\norm{\cdot}^\dagger\colon \Rmn \to \R_+$ is the dual norm associated with $\norm{\cdot}$, i.e., $\norm{\mX}^\dagger = \sup_{\norm{\mX'}\leq 1} \<\mX,\mX'>$ for all $\mX \in \Rmn$. The assumption of gradient Lipschitz continuity is also widespread in the analysis of first-order optimization methods \citep{ghadimi2013stochastic,gower2019sgd,cutkosky2020momentum,horvath2023stochastic,gorbunov2020linearly}. It is important to highlight that while \Cref{eq:L} uses the dual norm $\norm{\cdot}^\dagger$ to measure the difference between the gradients, the variance in \Cref{eq:variance} is measured with respect to the Frobenius norm $\sqnf{\cdot}$, which is necessary to properly utilize the unbiasedness property of the stochastic gradient estimator $g(\cdot;\xi)$. Therefore, we need to provide a connection between these norms using the following inequality:
    \begin{aequation}\label{eq:norm}
    \norm{\mX}^\dagger \leq \rho\cdot\normf{\mX}
    \quad\text{for all}\;
    \mX \in \Rmn,
    \end{aequation}
    where $\rho > 0$ is a positive constant. Note that such a constant always exists due to the norm equivalence theorem, which always holds in the finite-dimensional space $\Rmn$. We recount $\rho$ for different norms $\norm{\cdot}$ in the appendix.

@@

\section{Appendix}
\subsection{Norms $\normfstar{\cdot}^\dagger$ and $\normftwo{\cdot}^\dagger$}
Here we provide the derivation of these norms and plot them in the case of $2\times2$ matrices.
\subsection{Updates derivations}

>>>>

{\bf Non-Euclidean norm setting and Lipschitz continuous gradient.}
    We assume that matrix space $\Rmn$ is equipped with a norm $\norm{\cdot}\colon \Rmn \to \R_+$, which possibly does not coincide with the Frobenius, norm $\normf{\cdot}$. In addition, we assume that the gradient $\nabla f(\cdot)$ is Lipschitz continuous with respect to the norm $\norm{\cdot}$, that is, the following inequality holds:
    \begin{aequation}\label{eq:L}
    \norm{\nabla f(\mX) - \nabla f(\mX')}^\dagger \leq L \norm{\mX-\mX'}
    \quad\text{for all}\;
    \mX, \mX' \in \Rmn,
    \end{aequation}
    where $L > 0$ is the gradient Lipschitz constant, and $\norm{\cdot}^\dagger\colon \Rmn \to \R_+$ is the dual norm associated with $\norm{\cdot}$, i.e., $\norm{\mX}^\dagger = \sup_{\norm{\mX'}\leq 1} \<\mX,\mX'>$ for all $\mX \in \Rmn$.

@@

\section{Appendix}
\subsection{Formal assumptions}
{\bf Stochastic gradient estimator.}
    We assume access to a stochastic estimator $g(\cdot\,;\xi)\colon \Rmn \to \Rmn$ of the gradient $\nabla f(\cdot)$, where $\xi \sim \cD$ is a random variable sampled from a probability distribution $\cD$. We assume that the stochastic gradient estimator $g(\cdot;\xi)$ is unbiased and has bounded variance, that is, the following relations hold:
    \begin{aequation}\label{eq:variance}
    \Ed{\xi \sim \cD}{g(\mX;\xi)} = \nabla f(\mX)
    \quad\text{and}\quad
    \Ed{\xi \sim \cD}{\sqnf{g(\mX;\xi) - \nabla f(\mX)}} \leq \sigma^2
    \quad\text{for all}\;
    \mX \in \Rmn,
    \end{aequation}
    where $\sigma > 0$ is a positive variance parameter, and $\normf{\cdot}$ is the standard Euclidean, i.e. Frobenius, norm induced by the inner product $\<\cdot,\cdot>$, i.e., $\normf{\mX} = \sqrt{\<\mX,\mX>}=\sqrt{\tr(\mX^\top \mX)}$. These assumptions have been widely adopted for the analysis of many stochastic gradient optimization algorithms \citep{ghadimi2013stochastic,ghadimi2016accelerated,cutkosky2020momentum,sun2023momentum,horvath2023stochastic,gorbunov2020linearly}.

    It is important to highlight that while \Cref{eq:L} uses the dual norm $\norm{\cdot}^\dagger$ to measure the difference between the gradients, the variance in \Cref{eq:variance} is measured with respect to the Frobenius norm $\sqnf{\cdot}$, which is necessary to properly utilize the unbiasedness property of the stochastic gradient estimator $g(\cdot;\xi)$. Therefore, we need to provide a connection between these norms using the following inequality:
    \begin{aequation}\label{eq:norm}
    \norm{\mX}^\dagger \leq \rho\cdot\normf{\mX}
    \quad\text{for all}\;
    \mX \in \Rmn,
    \end{aequation}
    where $\rho > 0$ is a positive constant. Note that such a constant always exists due to the norm equivalence theorem, which always holds in the finite-dimensional space $\Rmn$. We recount $\rho$ for different norms $\norm{\cdot}$ in the appendix.

\subsection{Norms $\normfstar{\cdot}^\dagger$ and $\normftwo{\cdot}^\dagger$}
Here we provide the derivation of these norms and plot them in the case of $2\times2$ matrices.
\subsection{Updates derivations}
